---
title: "FlexDotPlot_Tutorial"
author: "Simon Leonard"
date: "22/02/01"
output:
  html_document:
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
fig_width: 8
fig_height: 20
vignette: >
  %\VignetteIndexEntry{FlexDotPlot_Tutorial}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = TRUE,
  fig.width = 16, fig.height = 6
)
```

## Package installation
```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("Simon-Leonard/FlexDotPlot")

```


## Package Loading
```{r message=FALSE}
library(FlexDotPlot)
```

***

## Run the dot_plot function

### 1. Example dataset {.tabset}

#### Command
```{r}
data(PBMC3K_example_data)
```


#### Input overview (head)
```{r}
head(PBMC3K_example_data)
```

#### Input overview (summary)
```{r}
summary(PBMC3K_example_data)
```



***

### 2. Choose variables to plot {.tabset}

#### Use the size parameters 

Size can be set with a column name using the size_var argument. Size scale can be adjusted with the shape.scale argument
```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, size_var = "pct.exp", shape.scale = 8)
```


A column index or a vector can also be provided 
```{r, eval=F}
# Same results
dot_plot(data.to.plot = PBMC3K_example_data, size_var = 3, shape.scale = 8) # Third column = pct.exp
dot_plot(data.to.plot = PBMC3K_example_data, size_var = PBMC3K_example_data$pct.exp, shape.scale = 8)
```

Comment : The variable controlling the shape size has to be numeric
```{r, eval=F}
class(PBMC3K_example_data$pval_symb)
dot_plot(data.to.plot = PBMC3K_example_data, size_var ="pval_symb") # Error ; size_var column has to be numeric
```

##### Additional size parameters

Scale the size by size or radius with the scale.by argument (default = "radius")
Set lower and upper limits for scaling with scale.min and scale.max arguments, use NA for default values.

***

#### Use the color parameters 

Color can be set with a column name using the col_var argument.
```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, col_var = "pct.exp", shape.scale = 20)
```


A column index or a vector can also be provided 
```{r, eval=F}
# Same results
dot_plot(data.to.plot = PBMC3K_example_data, col_var = 3, shape.scale = 20) # Third column = pct.exp
dot_plot(data.to.plot = PBMC3K_example_data, col_var = PBMC3K_example_data$pct.exp, shape.scale = 20)
```



Comment : A discrete variable can also be used to set the shape color
```{r}
class(PBMC3K_example_data$pval_symb)
dot_plot(data.to.plot = PBMC3K_example_data, col_var ="pval_symb", shape.scale=20) 
```


##### Use custom colors

Custom colors can be used with the cols.use parameter
```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, col_var = "pct.exp", cols.use = c("yellow","blue"), shape.scale = 20)
```

#### Use the shape parameters 

Variable shape can be set with a column name using the shape_var argument.
```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, shape_var = "pct.exp", shape.scale = 8)
```


A vector can also be provided 
```{r, eval=F}
# Same results
dot_plot(data.to.plot = PBMC3K_example_data, shape_var = PBMC3K_example_data$pct.exp, shape.scale = 8)
```



Comment : A discrete variable can also be used to set the shape
```{r}
class(PBMC3K_example_data$pval_symb)
dot_plot(data.to.plot = PBMC3K_example_data, shape_var ="pval_symb", shape.scale=20) 
```


You can choose the shape types with the shape_use argument and unicode characters
```{r}
dot_plot(data.to.plot = PBMC3K_example_data, shape_var ="pval_symb", shape_use=c("\u2605","\u2736","\u25CF","\u2737","\u2726"),
         shape.scale=20) 
```

#### Add text on shapes 

Text can be set with a column name using the text_var argument. Text size can be adjusted with the text.size argument
```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, text_var = "pct.exp2", text.size = 3, shape.scale = 20)
```


A column index or a vector can also be provided 
```{r, eval=F}
# Same results
dot_plot(data.to.plot = PBMC3K_example_data, text_var = 4, , text.size = 3, shape.scale = 20)
dot_plot(data.to.plot = PBMC3K_example_data, text_var = PBMC3K_example_data$pct.exp2, , text.size = 3, shape.scale = 20)
```


***

### 3. Combine variables

Shape, size, colors and text can be provided simultaneously and independently

```{r message=FALSE, warning=FALSE, fig.height= 4}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pct.exp",
         col_var = "avg_logFC"
         )

dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= ifelse(PBMC3K_example_data$avg_logFC>0, "Up","Down"), 
         col_var = "avg_logFC", cols.use = c("yellow","blue")
         )
```

### 4. Add dendrograms to the plot

Variable used to compute x and y dendrogramms can be provided within the dend_x_var and dend_y_var respectively.
dist_method and hclust_methods arguments control distance method (default = euclidean) and hclust method (default = ward.D).

x dendrogramm only : 


```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pct.exp",
         col_var = "avg_logFC",
         dend_x_var = c("pct.exp","avg_logFC"),
         dist_method="euclidean", hclust_method="ward.D"
         )
```

x and y dendrogramms :


```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pct.exp",
         col_var = "avg_logFC",
         dend_x_var = c("pct.exp","avg_logFC"),
         dend_y_var = c("pct.exp","avg_logFC"),
         dist_method="euclidean", hclust_method="ward.D"
         )
```

***
### 5. Customize plot legends {.tabset}

#### Shape legend

Legend title can be modified with the size_legend, col_legend and shape_legend arguments.

```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8, size_legend="My size legend",
         shape_var= "pval_symb", shape_legend="My shape legend",
         col_var = "avg_logFC", col_legend="My col legend"
         )
```

If you don't want to plot the legend, set the plot.legend argument to FALSE.

```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pval_symb",
         col_var = "avg_logFC",
         plot.legend=F
         )
```

size.breaks.number argument define the number of shapes with different size to display in the legend

```{r}
dot_plot(data.to.plot = PBMC3K_example_data, size_var = "pct.exp", shape.scale = 8, size.breaks.number = 3)
```

You can also directly specify the values to get in the legend with the size.breaks.values argument

```{r}
dot_plot(data.to.plot = PBMC3K_example_data, size_var = "pct.exp", shape.scale = 8, size.breaks.values = c(0,0.25,0.5,0.75,1))
```

color.breaks.number argument define the number of labels for the color gradient legend.

```{r}
dot_plot(data.to.plot = PBMC3K_example_data, col_var = "pct.exp", shape.scale = 20, color.breaks.number = 3)
```

You can also directly specify the values to get in the legend with the color.breaks.values argument

```{r}
dot_plot(data.to.plot = PBMC3K_example_data, col_var = "pct.exp", shape.scale = 20, color.breaks.values = c(0,0.3,0.6,0.9))
```


#### Axis labels

Axis label positions can be controled with the x.lab.pos and y.lab.pos arguments.
x.lab.pos must be one of "bottom","top","both" or "none".
y.lab.pos must be one of "left","right","both"or "none".

```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pct.exp",
         col_var = "avg_logFC",
         x.lab.pos="bottom",
         y.lab.pos="left"
         )
```


You can rotate x axis labels with the x.lab.rot argument (default is TRUE). Set it to FALSE if you don't want to rotate x axis labels (It's not recommended here)


Plot can be transposed by setting the transpose argument to TRUE

```{r, fig.width=6, fig.height=12}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pct.exp",
         col_var = "avg_logFC",
         transpose=TRUE
         )
```

***
### 6. Plot background {.tabset}

#### Vertical shades
Vertical shades can be added with the vertical_coloring argument 

```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pct.exp",
         col_var = "avg_logFC",
         vertical_coloring=c(NA, "gray80")
         )
```

#### Horizontal shades

Horizontal shades can be added with the horizontal_coloring argument 

```{r message=FALSE, warning=FALSE}
dot_plot(data.to.plot = PBMC3K_example_data, 
         size_var = "pct.exp", shape.scale = 8,
         shape_var= "pct.exp",
         col_var = "avg_logFC",
         horizontal_coloring=c(NA, "gray80")
         )
```


### 7. Get the corresponding ggplot object

Set do.return to TRUE to get the corresponding ggplot object

```{r eval=FALSE}
ggplot_object <- dot_plot(data.to.plot = PBMC3K_example_data, 
                          size_var = "pct.exp", shape.scale = 8,
                          shape_var= "pct.exp",
                          col_var = "avg_logFC",
                          do.return = TRUE
                          )
```


## Run the Shiny app 


```{r eval=FALSE}
Shiny_dot_plot() 
```


## Reproduce the article figure 

Full script to generate dotplot input can be found at in the FlexDotPlot_CBMC_dataset vignette

```{r}
### Import data + create the exp.type column
data(CBMC8K_example_data)

### Dotplot
dotplot = dot_plot( 
  data.to.plot=CBMC8K_example_data, 
  size_var="RNA.avg.exp.scaled",col_var="ADT.avg.exp.scaled", text_var="ADT.pct.exp.sup.cutoff",
  shape_var="canonical_marker", shape_use = c("\u25CF","\u2737"),
  x.lab.pos="bottom",y.lab.pos="left", 
  cols.use=c("lightgrey","orange","red", "darkred"),size_legend="RNA", col_legend="ADT", shape_legend="Canonical marker ?", 
  shape.scale =12, text.size=3, 
  plot.legend = TRUE,x.lab.rot = TRUE, 
  size.breaks.number=4, color.breaks.number=4, shape.breaks.number=5
  , dend_x_var=c("RNA.avg.exp.scaled","ADT.avg.exp.scaled"), dend_y_var=c("RNA.avg.exp.scaled","ADT.avg.exp.scaled"), dist_method="euclidean",
  hclust_method="ward.D", do.return = TRUE)

 
```


## Session Info
```{r echo=FALSE}
sessionInfo()
```

