---
title: "Dotplot with cellphoneDB dataset"
author: "Simon Leonard"
date: "22/02/01"
output:
  html_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{FlexDotPlot_cellphoneDB_dataset}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

## Package + input Loading

```{r message=FALSE}
library(Seurat)
library(data.table)
library(ggplot2)
library(FlexDotPlot)
data("CellphoneDB_example_data")
```


### Generate dotplot : cellphonedb/original figure way

```{r}
my_palette <- colorRampPalette(c("#179493","#76c286", "#ebdc96", "#ec9173","#d0587e"), alpha=TRUE)(n=399)

ggplot(CellphoneDB_example_data,aes(x=clusters,y=pair)) +
  geom_point(aes(size=-log10(pvalue),color=mean)) +
  scale_color_gradientn('Log2 mean (Molecule 1, Molecule 2)', colors=my_palette) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.text=element_text(size=14, colour = "black"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size=12, colour = "black"),
        axis.title=element_blank(),
        panel.border = element_rect(size = 0.7, linetype = "solid", colour = "black"))

```


### Split each dotplot column to have one column for receptor and one other for ligand

```{r}
CellphoneDB_example_data=CellphoneDB_example_data[order(CellphoneDB_example_data$clusters, decreasing = F),]
CellphoneDB_example_data$clusters=as.character(CellphoneDB_example_data$clusters)

res=data.frame()
for (i in 1:nrow(CellphoneDB_example_data)){
  res1=CellphoneDB_example_data[i,c(1:4,seq(5,9,2),11)]
  res2=CellphoneDB_example_data[i,c(1:4,seq(6,10,2),11)]
  
  colnames(res1)[4:7]=colnames(res2)[4:7]=c("log2mean_pair","mean_gene","pct_gene","log2mean_gene")
  res1$clusters=paste(res1$clusters,"1",sep="_")
  res2$clusters=paste(res2$clusters,"2",sep="_")
  
  res=rbind(res,res1,res2)
}

res$clusters=factor(res$clusters, levels = unique(res$clusters))
res$round_mean_gene=round(res$mean_gene, 1)

```

### Dotplot with FlexDotPlot

```{r message=FALSE, warning=FALSE, fig.width= 6, fig.height=6}
dot_plot(res, shape_var = "pct_gene", col_var = 'log10pval', 
         # text_var = "round_mean_gene",
         shape.scale = 14,
         color.breaks.values = c(0,1,2,3), shape.breaks.number = 5, size.breaks.number = 5,
         x.lab.pos = "bottom",vertical_coloring = c(NA,NA,"grey","grey"),
         col_legend = "-log10(pval)", cols.use = c("white","blue"),
         text.size = 3,
         transpose = T)
```


## Session Info
```{r}
sessionInfo()
```



## Generate  CellphoneDB_example_data from raw scRNAseq data

### Import data and create seurat object

Data from : <https://doi.org/10.1016/j.ccell.2021.02.015>

Data availability : <https://singlecell.broadinstitute.org/single_cell/study/SCP1288/tumor-and-immune-reprogramming-during-immunotherapy-in-advanced-renal-cell-carcinoma#study-summary>

```{r message=FALSE, warning=FALSE, eval=FALSE}

count=fread("E:/Analyses_10X_ccRCC_Bi_2021/SCP1288/expression/ccRCC_scRNASeq_NormalizedCounts.txt", sep="\t", header=T, data.table = F)
rownames(count)=count$GENE
count$GENE=NULL

meta=read.table("E:/Analyses_10X_ccRCC_Bi_2021/SCP1288/cluster/Final_SCP_ClusterFile.txt", header=T, sep="\t", row.names = 1)
meta=meta[-1,]
colnames(meta)[1:2]=c("UMAP_1","UMAP_2")
meta$UMAP_1=as.numeric(meta$UMAP_1); meta$UMAP_2=as.numeric(meta$UMAP_2)

seurat=CreateSeuratObject(counts = as.matrix(count), meta.data = meta)
seurat[["umap"]]=CreateDimReducObject(embeddings = as.matrix(meta[,c("UMAP_1","UMAP_2")]), key="umap_")
DimPlot(seurat, group.by = "Lineage")

rm(count, meta)

```

### Create custom clusters to aggregate all CD8+ T cells and all TAMs

```{r, eval=FALSE}
seurat$cluster=seurat$FinalCellType
seurat$cluster[grep("CD8\\+ T cell", seurat$cluster)]="CD8+ T cell"
seurat$cluster[grep("TAM", seurat$cluster)]="TAM"

# Subset seurat to keep cells used in cellphoneDB
Idents(seurat)="cluster"
subseurat=subset(seurat, idents = c("CD8+ T cell", "TAM","TP1","TP2"))
```

### Generate/write cellphonedb inputs

```{r eval=FALSE}

write.table(seurat@assays$RNA@data, "E:/Analyses_10X_ccRCC_Bi_2021/cellphonedb/21_06_10_cellphonedb_count_CD8_TAM_Tum.txt", sep="\t", quote=F)

# generating meta file
meta_data <- cbind(rownames(subseurat@meta.data), subseurat@meta.data[,"cluster", drop=F])   #####  cluster is the user's specific cluster column
write.table(meta_data, "E:/Analyses_10X_ccRCC_Bi_2021/cellphonedb/21_06_10_cellphonedb_meta_CD8_TAM_Tum.txt", sep="\t", quote=F, row.names=F)

```

## Run cellphoneDB

```{python, eval=FALSE}
cellphonedb method statistical_analysis --counts-data gene_name 21_06_10_cellphonedb_meta_CD8_TAM_Tum.txt 21_06_10_cellphonedb_count_CD8_TAM_Tum.txt
```

## Process cellphoneDB outputs

Adapted from dotplot function in cellphonedb : <https://github.com/Teichlab/cellphonedb/blob/master/cellphonedb/src/plotters/R/plot_dot_by_column_name.R>

### Input/parameters

```{r, eval=FALSE}
dir="E:/Analyses_10X_ccRCC_Bi_2021/cellphonedb/"
selected_rows = read.delim(paste(dir,"dotplot_rows.txt",sep=""), header=F)[,1]
selected_columns = read.delim(paste(dir,"dotplot_cols.txt",sep=""), header=F)[,1]
means_path = paste(dir,'/out/means.txt', sep="")
pvalues_path = paste(dir,'/out/pvalues.txt', sep="")
means_separator = '\t'
pvalues_separator = '\t'
```

### Process cellphoneDB outputs

```{r, eval=FALSE}
all_pval = read.table(pvalues_path, header=T, stringsAsFactors = F, sep=means_separator, comment.char = '', check.names=F)
all_means = read.table(means_path, header=T, stringsAsFactors = F, sep=pvalues_separator, comment.char = '', check.names=F)

intr_pairs = all_pval$interacting_pair
all_pval = all_pval[,-c(1:11)]
all_means = all_means[,-c(1:11)]

if(is.null(selected_rows)){
  selected_rows = intr_pairs
}

if(is.null(selected_columns)){
  selected_columns = colnames(all_pval)
}

sel_pval = all_pval[match(selected_rows, intr_pairs), selected_columns]
sel_means = all_means[match(selected_rows, intr_pairs), selected_columns]

df_names = expand.grid(selected_rows, selected_columns)
pval = unlist(sel_pval)
pval[pval==0] = 0.0009
plot.data = cbind(df_names,pval)
pr = unlist(as.data.frame(sel_means))
pr[pr==0] = 1
plot.data = cbind(plot.data,log2(pr))
colnames(plot.data) = c('pair', 'clusters', 'pvalue', 'mean')
```

### Process dotplot input to get same dotplot as in the original paper-

```{r, eval=FALSE}
# Reverse some interactions according to original figure
int_to_revers=c("IFNG_Type II IFNR", "CCL4_CCR5", "CCL3_CCR1", "HLA-F_LILRB1", "TIGIT_NECTIN2", "PDCD1_PDCD1LG2")

pair.levels=levels(plot.data$pair)
clus.levels=levels(plot.data$clusters)

plot.data$pair=as.character(plot.data$pair)
plot.data$clusters=as.character(plot.data$clusters)
for (i in 1:nrow(plot.data)){
  if (plot.data[i,"pair"] %in% int_to_revers){
    x=plot.data[i,"pair"]
    plot.data[i,"pair"]=paste(substring(x,as.numeric(gregexpr("_",x))+1),substring(x,1,as.numeric(gregexpr("_",x))-1), sep="_")
    
    x=plot.data[i,"clusters"]
    plot.data[i,"clusters"]=paste(substring(x,as.numeric(gregexpr("\\|",x))+1),substring(x,1,as.numeric(gregexpr("\\|",x))-1), sep="|")
  }
}
plot.data$pair=factor(plot.data$pair, levels = unique(plot.data$pair))
plot.data$clusters=factor(plot.data$clusters, levels = clus.levels)

plot.data=plot.data[plot.data$clusters %in% levels(plot.data$clusters)[1:5],]

```

### Add some expression metrics for each pair

```{r, eval=FALSE}
plot.data$mean1=NA
plot.data$mean2=NA
plot.data$pct1=NA
plot.data$pct2=NA
for (i in 1:nrow(plot.data)){
  x=substring(plot.data[i,"pair"],1,as.numeric(gregexpr("_",plot.data[i,"pair"]))-1)
  y=substring(plot.data[i,"clusters"],1,as.numeric(gregexpr("\\|",plot.data[i,"clusters"]))-1)
  if(x %in% rownames(subseurat)){
    fetch=FetchData(subseurat, x, cells=WhichCells(subseurat, idents=y))[,1]
  }else if (x=="Type II IFNR"){
    fetch=unlist(FetchData(subseurat, c("IFNGR1", "IFNGR2"), cells=WhichCells(subseurat, idents=y)))
  }
  plot.data$mean1[i]=mean(fetch)
  plot.data$pct1[i]=round(sum(fetch>0)/length(fetch)*100)
  
  x=substring(plot.data[i,"pair"],as.numeric(gregexpr("_",plot.data[i,"pair"]))+1)
  y=substring(plot.data[i,"clusters"],as.numeric(gregexpr("\\|",plot.data[i,"clusters"]))+1)
  if(x %in% rownames(subseurat)){
    fetch=FetchData(subseurat, x, cells=WhichCells(subseurat, idents=y))[,1]
  }else if (x=="Type II IFNR"){
    fetch=unlist(FetchData(subseurat, c("IFNGR1", "IFNGR2"), cells=WhichCells(subseurat, idents=y)))
  }
  plot.data$mean2[i]=mean(fetch)
  plot.data$pct2[i]=round(sum(fetch>0)/length(fetch)*100)
}

plot.data$log2mean1=log2(plot.data$mean1)
plot.data$log2mean2=log2(plot.data$mean2)
plot.data$log10pval=-log10(plot.data$pvalue)
plot.data$clusters=factor(plot.data$clusters, levels=intersect(levels(plot.data$clusters), plot.data$clusters))

```
